### 2016-11-01 Paper_v8.tex
added extra ZEB refs

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
Lu \emph{et al.} modeled the role that micro-RNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa}.
Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.

@@

NF-$\kappa$B may also play an essential role of in the epithelial transformation.
NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \cite{Wu:2009cr}.  This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \cite{Sullivan:2009bh}.
Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
In addition, other important regulatory proteins such as ZEB1 should be added to the next generation model.
ZEB1 can activate the expression of mesenchymal genes such as collagens, smooth muscle actin and myosin, Vimentin while simultaneously contributing
to the repression of epithelial genes \cite{Liu:2008aa}.
Thus, ZEB1 can act in parallel to SNAIL and thus be an alternative to the canonical signaling paradigm presented here.
Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs,
for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and Lu et al have suggested that
EMT is driven by a miR-34/SNAIL and miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa}.
Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.

@@

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig1_coarse_network.pdf}
\caption{Model connectivity recreates the core architecture during EMT. The EMT network contains 97 nodes (proteins, mRNA, and genes) interconnected by 169 interactions.
Central to EMT induction, activation of the MAPK cascade occurs through TGF-$\beta$1/2 binding which activates the AP-1/Sp1 transcriptional axis. AP-1/Sp1 drives an autocrine response of TGF-$\beta$3, which activates the Smad cascade, leading to phenotypic change. Conversely, VEGF-A binding can stabalize an epithelial phenotype through NFAT activation.
Downstream activation of $\beta$-catenin signaling due to E-cadherin loss and GSK3 inactivation of $\beta$-cateinin confinement is critical to the complete activation of the EMT program.
The complete list of molecular interactions that comprise the model is given in the supplement.}\label{fg:F1}
\end{figure}

>>>>

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
Several studies on ZEB signaling have modeled the role that micro-RNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and/or miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.

@@

NF-$\kappa$B may also play an essential role of in the epithelial transformation.
NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \cite{Wu:2009cr}.  This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \cite{Sullivan:2009bh}.
Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
In addition, other important regulatory proteins such as ZEB1 should be added to the next generation model.
ZEB1 can activate the expression of mesenchymal genes such as collagens, smooth muscle actin and myosin, Vimentin while simultaneously contributing
to the repression of epithelial genes \cite{Liu:2008aa}.
Thus, ZEB1 can act in parallel to SNAIL and thus be an alternative to the canonical signaling paradigm presented here.
Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs,
for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and have suggested that
the EMT process is driven by a miR-34/SNAIL and/or miR-200/ZEB feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.

@@

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig1_coarse_network.pdf}
\caption{Model connectivity recreates the core architecture during EMT. The EMT network contains 97 nodes (proteins, mRNA, and genes) interconnected by 169 interactions.
Central to EMT induction, activation of the MAPK cascade occurs through TGF-$\beta$1/2 binding which activates the AP-1/Sp1 transcriptional axis. AP-1/Sp1 drives an autocrine response of TGF-$\beta$3, which activates the Smad cascade, leading to phenotypic change. Conversely, VEGF-A binding can stabilize an epithelial phenotype through NFAT activation.
Downstream activation of $\beta$-catenin signaling due to E-cadherin loss and GSK3 inactivation of $\beta$-catenin confinement is critical to the complete activation of the EMT program.
The complete list of molecular interactions that comprise the model is given in the supplement.}\label{fg:F1}
\end{figure}
